Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-28 PAPER.tex
removed references to 8 independent modes

This paper is organized into two main sections. In Section \ref{sec:Themes} we introduce the three themes of our focus, using a 
toy model to develop intuition for each one. In Section \ref{sec:CaseStudy} we present a case study into each theme using data 
from the PAPER-64 array, highlighting key changes from \citet{ali_et_al2015}, henceforth known as P64, which have led to a 
revised PAPER-64 power spectrum result (\cc{cite retraction or kolopanis paper}). We conclude in Section \ref{sec:Con}.

@@

Because fringe-rate filtering is analogous to averaging in time, it comes at the cost of reducing the total number of independent 
samples in the data. To mimic this filter, we average every four time integrations of our toy model dataset together, yielding 
$25$ independent samples in time (Figure \ref{fig:toy_sigloss5}, left). We choose these numbers so that the total number of 
independent samples is similar to the number of frequency channels --- therefore, our matrices will be full rank.\footnote{If 
instead we average every $20$ samples together, yielding only $5$ independent samples in time, we are left with a rank 
deficient covariance matrix. In this case... \cc{fill in}}

The resulting eigenspectrum as compared to the green curve (FG + EoR) in Figure \ref{fig:toy_sigloss2} is shown in Figure 
\ref{fig:toy_sigloss5} (right plot). The spectrum, in the case of fringe-rate filtering (dashed line), falls more steeply, especially for 
the last few eigenmodes. This is because we have fewer independent modes --- \cc{need help filling out the rest here... and 
also citing galaxy lit}.

@@

\noindent of $(22.4$ mK$)^{2}$ in the range $0.15 < k < 0.5$ h Mpc$^{-1}$ at $z = 8.4$. %The revision of this limit stems 
mostly from previously underestimated signal loss and underestimated error bars \cc{cite retraction paper}, both of which we 
address in the following sections.

@@

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method used in P64, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One 
specific change from P64 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss, likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the 
same level if different $21$ cm simulations were used remains an open question.}

@@

To conclude this method, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal 
loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted 
power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and 
grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it 
is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). 
Post-correction, the power spectrum values blow up to be higher than both the theory and unweighted power spectrum.

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of 
independent samples in PAPER-64. PAPER-64 only has $\sim$$8$ independent modes after fringe-rate filtering while our 
power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe 
real fluctuations beyond the first eight. Hence, there is a substantial down-weighting of other modes, a large loss that the signal 
loss transfer function captures and corrects for. We elaborate on this point in the next section, as well as investigate alternate 
weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-
weighting contaminants with minimizing the loss of EoR.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the 
eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes 
down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse 
covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss 
correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting 
becomes more aggressive (grey curve). In other words, more `weak' (EoR-dominated) fluctuations are being overfit and 
subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, 
increases with the number of down-weighted modes (black curve). As discussed previously, we are limited in our power 
spectrum accuracy by the number of independent samples in our dataset and our errors dramatically increase as we surpass 
the $\sim8$ independent samples that we have in fringe-rate filtered PAPER-64 data \cc{elaborate?}.

Optimistically, we expect there to be a `sweet spot' as we dial our regularization knob; a level of regularization where weighting 
is beneficial compared to not weighting (blue dashed line). We see in Figure \ref{fig:sigloss_modeloop} that this occurs when 
only the strongest $\sim3$-$4$ eigenmodes are down-weighted, though the improvement from the unweighted case is not 
significant.

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure 
\ref{fig:ps1_data}. Again, black and grey points correspond to positive and negative power spectrum values respectively, with 
$2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of 
noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, 
a higher limit than P64 by a factor of $\sim7$ in mK (though we only use one baseline type in our analysis as opposed to $3$). 
\cc{ref kolopanis limit instead later}

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and grey points correspond to positive and 
negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted 
power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical $2\sigma$ noise level prediction based on 
observational parameters. This power spectrum result differs from P64 in that it only uses data from one type of baseline ($30$ m 
East/West baselines) instead of three. Major differences from previously published results stem from revisions regarding signal 
loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at 
$k=0.34$ $h$Mpc$^{-1}$, a higher limit than P64 by a factor of $\sim7$ in mK \cc{ref kolopanis limit instead later}.}
	\label{fig:ps1_data}
\end{figure*}

In this section we have shown three simple ways of regularizing $\hat{\textbf{C}}$ to minimize signal loss using PAPER-64 
data. There are many other weighting schemes that we leave for consideration in future work. For example, one could estimate 
$\hat{\textbf{C}}$ using information from different subsets of baselines. For redundant arrays this could mean calculating $
\hat{\textbf{C}}$ from a different but similar baseline type, such as the $\sim30$ m diagonal PAPER baselines (instead of the 
horizontal E/W ones). Alternately, covariances could be estimated from all other baselines except the two being cross-multiplied 
when forming a power spectrum estimate. This method was used in \citet{parsons_et_al2014} in order to avoid suppressing the 
$21$ cm signal, and it's worth noting that the PAPER-32 results are likely safe from the issue of signal loss underestimation 
because of this very reason (however, they are affected by the error estimation issues described in Section \ref{sec:Error}, so 
we also regard those results as suspect and superseded by those of \cc{cite kolopanis}).

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $144$ K, yielding $T_{sys} = 431$ K at $150$ MHz. 
This is lower than the $T_{sys}$ of $500$ K used in P64 because of several small miscalculation errors that were 
identified\footnote{such as missing a square root in going from a variance to a standard deviation.}.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum 
estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root 
because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the 
analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The updated limits in \cc{cite kolopanis paper} 
use three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a 
fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time 
using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is 
taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we 
divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, 
because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of 
$N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ 
days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was 
the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects 
the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is:
\begin{equation}
N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}},
\end{equation}
\noindent where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines 
entering the group ($N_{bls}/N_{gps}$) and then as the square root of the number of cross-multiplied pairs ($\sqrt{N_{gps}^{2} - 
N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$ instead 
of $51$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

In P64, a second round of bootstrapping occurred over the time axis. A total of $400$ bootstraps were created in this step 
($N_{boot} = 400$), each comprised of randomly selected values sampled with replacement along the time axis. More 
specifically, each of these bootstraps contained the same number of values as the number of time integrations (which, at $\sim$
$700$, greatly exceeds the approximate number of independent samples after fringe-rate filtering, which is $\sim$$8$). 
Medians were then taken of the values in each bootstrap (with the appropriate median correction factor applied). Finally, power 
spectrum limits were computed by taking the mean and standard deviation over all the bootstraps. We emphasize again that in 
this previous analysis, the number of elements sampled per bootstrap ($N \sim $ number of times $\sim$ $700$) greatly 
exceeded the number of independent LST samples ($\sim$$8$), under-estimating errors. A random draw of $700$ 
measurements from this dataset has many repeated values, and the variance between hundreds ($N_{boot}$) of these random 
samples is smaller than the true underlying variance of the data.

>>>>

This paper is organized into two main sections. In Section \ref{sec:Themes} we introduce the three themes of our focus, using a 
toy model to develop intuition for each one. In Section \ref{sec:CaseStudy} we present a case study into each theme using data 
from the PAPER-64 array, highlighting key changes from \citet{ali_et_al2015}, henceforth known as P64, which have led to a 
revised PAPER-64 power spectrum result (Kolopanis et al., \textit{in prep}). We conclude in Section \ref{sec:Con}.

@@

Because fringe-rate filtering is analogous to averaging in time, it comes at the cost of reducing the total number of independent 
samples in the data. To mimic this filter, we average every four time integrations of our toy model dataset together, yielding 
$25$ independent samples in time (Figure \ref{fig:toy_sigloss5}, left). We choose these numbers so that the total number of 
independent samples is similar to the number of frequency channels --- therefore, our matrices will be full rank.
%\footnote{If instead we average every $20$ samples together, yielding only $5$ independent samples in time, we are left with a rank deficient covariance matrix. In this case... \cc{fill in}}

The resulting eigenspectrum as compared to the green curve (FG + EoR) in Figure \ref{fig:toy_sigloss2} is shown in Figure 
\ref{fig:toy_sigloss5} (right plot). The spectrum, in the case of fringe-rate filtering (dashed line), falls more steeply, especially for 
the last few eigenmodes. This is because we have fewer independent modes and therefore less information to characterize the eigenspectrum. Because we do a worse job characterizing, we have more signal loss. \cc{need help filling out the rest here... and 
also citing galaxy lit}.

@@

\noindent of $(22.4$ mK$)^{2}$ in the range $0.15 < k < 0.5$ h Mpc$^{-1}$ at $z = 8.4$. The revision of this limit (Kolopanis et al., \textit{in prep}) stems mostly from previously underestimated signal loss and underestimated error bars, both of which we 
address in the following sections.

@@

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method used in P64, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One 
specific change from P64 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss (by a factor of $\sim$$10$), likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the same level if different $21$ cm simulations were used remains an open question.}

@@

To conclude this method, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal 
loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted 
power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and 
grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it 
is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). 
Post-correction, the power spectrum errors blow up to be higher than both the theory and unweighted power spectrum, a consequence of not being able to characterize covariances well using the time-averaged fringe-rate filtered data (there is less independent information to use). We elaborate on this point in the next section, as well as investigate alternate 
weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-
weighting contaminants with minimizing the loss of EoR.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the 
eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes 
down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse 
covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss 
correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting 
becomes more aggressive (grey curve). In other words, more `weak' (EoR-dominated) fluctuations are being overfit and 
subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, 
increases with the number of down-weighted modes (black curve). Because our data is fringe-rate filtered, we have less information to characterize the covariance matrix and therefore do a worse job estimating it. Hence, the more modes we use in down-weighting, the greater the error we have in estimating the power spectrum.

Optimistically, we expect there to be a `sweet spot' as we dial our regularization knob; a level of regularization where weighting 
is beneficial compared to not weighting (blue dashed line). In other words, we would like a weighting scheme that down-weights eigenmodes that describe foreground modes, but not EoR modes. We see in Figure \ref{fig:sigloss_modeloop} that this occurs when 
only the strongest $\sim3$-$4$ eigenmodes are down-weighted, though the improvement from the unweighted case is not 
significant.

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure 
\ref{fig:ps1_data}. Again, black and grey points correspond to positive and negative power spectrum values respectively, with 
$2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of 
noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, 
a higher limit than P64 by a factor of $\sim7$ in mK (though we only use one baseline type in our analysis as opposed to $3$). 
\cc{replace with kolopanis limit instead later}

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and grey points correspond to positive and 
negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted 
power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical $2\sigma$ noise level prediction based on 
observational parameters. This power spectrum result differs from P64 in that it only uses data from one type of baseline ($30$ m 
East/West baselines) instead of three. Major differences from previously published results stem from revisions regarding signal 
loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at 
$k=0.34$ $h$Mpc$^{-1}$, a higher limit than P64 by a factor of $\sim7$ in mK \cc{replace with kolopanis limit later}.}
	\label{fig:ps1_data}
\end{figure*}

In this section we have shown three simple ways of regularizing $\hat{\textbf{C}}$ to minimize signal loss using PAPER-64 
data. There are many other weighting schemes that we leave for consideration in future work. For example, one could estimate 
$\hat{\textbf{C}}$ using information from different subsets of baselines. For redundant arrays this could mean calculating $
\hat{\textbf{C}}$ from a different but similar baseline type, such as the $\sim30$ m diagonal PAPER baselines (instead of the 
horizontal E/W ones). Alternately, covariances could be estimated from all other baselines except the two being cross-multiplied 
when forming a power spectrum estimate. This method was used in \citet{parsons_et_al2014} in order to avoid suppressing the 
$21$ cm signal, and it's worth noting that the PAPER-32 results are likely safe from the issue of signal loss underestimation 
because of this very reason (however, they are affected by the error estimation issues described in Section \ref{sec:Error}, so 
we also regard those results as suspect and superseded by those of Kolopanis et al., \textit{in prep}.

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $144$ K, yielding $T_{sys} = 431$ K at $150$ MHz. 
This is lower than the $T_{sys}$ of $500$ K used in P64 because of several small miscalculation errors that were 
identified\footnote{such as missing a square root in going from a variance to a standard deviation}.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum 
estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root 
because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the 
analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The updated limits in Kolopanis et al., \textit{in prep}
use three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a 
fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time 
using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is 
taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we 
divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, 
because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of 
$N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ 
days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was 
the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects 
the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is:
\begin{equation}
N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}},
\end{equation}
\noindent where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines 
entering the group ($N_{bls}/N_{gps}$) and then as the square root of the number of cross-multiplied pairs ($\sqrt{N_{gps}^{2} - 
N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$ instead 
of $51$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

In P64, a second round of bootstrapping occurred over the time axis. A total of $400$ bootstraps were created in this step 
($N_{boot} = 400$), each comprised of randomly selected values sampled with replacement along the time axis. More 
specifically, each of these bootstraps contained the same number of values as the number of time integrations (which, at $\sim$
$700$, which we expect to greatly exceed the approximate number of independent samples after fringe-rate filtering.
Medians were then taken of the values in each bootstrap (with the appropriate median correction factor applied). Finally, power 
spectrum limits were computed by taking the mean and standard deviation over all the bootstraps. We emphasize again that in 
this previous analysis, the number of elements sampled per bootstrap greatly 
exceeded the number of independent LST samples, under-estimating errors. A random draw of $700$ 
measurements from this dataset has many repeated values, and the variance between hundreds ($N_{boot}$) of these random 
samples is smaller than the true underlying variance of the data.
